---
title: "Bulk_rna_seq"
author: "Muskan"
date: "2026-05-09"
output: html_document
---
#chunk1: fixing directory
```{r}
knitr::opts_chunk$set(
  echo = TRUE
)
knitr::opts_knit$set(root.dir = "E:/")
getwd()
```

#chunk2a:loading data and metadata and keeping just protein coding genes
```{r}
library(tidyverse)
setwd("E:/seq_bulk/1_SRR_to_IGV_bash_script")
getwd()
metadata <- read.csv("metadata_subset.csv")
merged <- read.csv("ileum_raw_counts.csv", check.names = FALSE, row.names = 1)

metadata
merged

dim(metadata)
dim(merged)

# Ensure gene names are clean
rownames(merged) <- make.unique(rownames(merged))
############################################################
# READ PROTEIN CODING GENE FILE
############################################################
setwd("E:/seq_bulk/2_gtf_to_protein_coding_genes")
pc_df <- read.delim(
  "pc_gene.txt",
  header = TRUE,
  sep = "\t",
  stringsAsFactors = FALSE
)
protein_coding_genes <- unique(
  pc_df$gene_name
)
merged <- merged[
  rownames(merged) %in% protein_coding_genes,
]
dim(merged)

intersect(colnames(merged), metadata$Sample)
metadata <- metadata[metadata$Sample %in% colnames(merged), ]
metadata <- metadata[match(colnames(merged), metadata$Sample), ]
all(metadata$Sample == colnames(merged))
```

#chunk2b: To find percenatge missing_genes_per_samples
```{r}
merged[is.na(merged)] <- 0
merged[!is.finite(as.matrix(merged))] <- 0

sample_missing_pct <- colSums(merged == 0) / nrow(merged)
missing_df <- data.frame(Sample = colnames(merged),              Missing_Percentage = sample_missing_pct[colnames(merged)])
write.csv(missing_df, "Percentage_of_missing_genes_per_samples.csv")

png("output/Histogram_percentage_of_missing_values_per_samples.png", width = 10, height = 8, units = 'in', res = 300)
hist(missing_df$Missing_Percentage, main = "Missing Genes per Sample", xlab = "Fraction Missing")
dev.off()

dim(merged)
write.csv(merged, "output/raw_merged.csv")
```

#chunk3: Batch correction using limma
```{r}
# Reorder to match sample order in merged ---
metadata <- metadata[match(colnames(merged), metadata$Sample), ]

# Sanity check (must be TRUE) ---
stopifnot(all(metadata$Sample == colnames(merged)))

# Prepare design matrix & batch variable ---
metadata$Batch <- as.factor(metadata$Batch)

mod <- model.matrix(~ Condition, data = metadata)

library(limma)
# Perform batch correction ---
corrected_data_limma_bc <- removeBatchEffect(as.matrix(merged),
                                             batch = metadata$Batch,
                                             design = mod)
```

#chunk4: Drop confounding Batches and then removing batch effect using removebatcheffect function of limma
```{r}
print(table(metadata$Batch, metadata$Condition))

exclude <- c("GSE183620", "GSE193677")
metadata$Batch <- droplevels(metadata$Batch)
metadata <- metadata[!metadata$Batch %in% exclude, ]
merged   <- merged[, colnames(merged) %in% metadata$Sample]

metadata <- metadata[match(colnames(merged), metadata$Sample), ]
metadata$Batch <- as.factor(metadata$Batch)

mod <- model.matrix(~ Condition, data = metadata)

corrected_data_limma_bc <- removeBatchEffect(as.matrix(merged),
                                             batch = metadata$Batch,
                                             design = mod)
write.csv(corrected_data_limma_bc, "output/limma/limma_merged.csv")
```

#chunk5: PCA plot before and after batch correction using limma.
```{r}
library(ggplot2)

# --- PCA Before Batch Correction ---
#remove genes that shows zero variance across samples.
merged <- merged[apply(merged, 1, var) > 0, ]
pca_before <- prcomp(t(merged), scale. = TRUE)
pca_before_df <- data.frame(pca_before$x, Condition = metadata$Condition, Batch = metadata$Batch)
var_explained_before <- round(100 * (pca_before$sdev^2 / sum(pca_before$sdev^2)), 2)

# --- PCA After Batch Correction ---
corrected_data_limma_bc <- corrected_data_limma_bc[apply(corrected_data_limma_bc, 1, var) > 0, ]
pca_after <- prcomp(t(corrected_data_limma_bc), scale. = TRUE)
pca_after_df <- data.frame(pca_after$x, Condition = metadata$Condition, Batch = metadata$Batch)
var_explained_after <- round(100 * (pca_after$sdev^2 / sum(pca_after$sdev^2)), 2)

# --- Plot Function ---
plot_pca <- function(pca_df, var_exp, title) {
  ggplot(pca_df, aes(x = PC1, y = PC2, color = Condition)) +
    geom_point(size = 3, alpha = 0.8) +
    theme_minimal(base_size = 14) +
    labs(
      title = title,
      x = paste0("PC1 (", var_exp[1], "%)"),
      y = paste0("PC2 (", var_exp[2], "%)"),
      color = "Condition"
    ) +
    theme(
      legend.position = "right",
      plot.title = element_text(face = "bold", hjust = 0.5)
    )
}

plot_pcab <- function(pca_df, var_exp, title) {
  ggplot(pca_df, aes(x = PC1, y = PC2, color = Batch)) +
    geom_point(size = 3, alpha = 0.8) +
    theme_minimal(base_size = 14) +
    labs(
      title = title,
      x = paste0("PC1 (", var_exp[1], "%)"),
      y = paste0("PC2 (", var_exp[2], "%)"),
      color = "Batch"
    ) +
    theme(
      legend.position = "right",
      plot.title = element_text(face = "bold", hjust = 0.5)
    )
}

p3 <- plot_pca(pca_before_df, var_explained_before, "PCA Global Before - by Condition")
p4 <- plot_pca(pca_after_df, var_explained_after, "PCA Global Final - by Condition")
p2 <- plot_pcab(pca_before_df, var_explained_before, "PCA Global Before - by Batch")
p1 <- plot_pcab(pca_after_df, var_explained_after, "PCA Global Final - by Batch")

ggsave("output/limma/PCA/PCA_before_Condition.png", p3, width = 8, height = 6, dpi = 300)
ggsave("output/limma/PCA/PCA_after_Condition.png", p4, width = 8, height = 6, dpi = 300)
ggsave("output/limma/PCA/YESPCA_before_Batch.png", p2, width = 8, height = 6, dpi = 300)
ggsave("output/Limma/PCA/PCA_after_Batch.png", p1, width = 8, height = 6, dpi = 300)
```

#chunk6: Use limma i.e batch corrected gene matrix file.
```{r}
library(limma)
library(edgeR)
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(ggplot2)
library(tibble)
library(readr)
library(ReactomePA)
```
```{r}
ileum_data <- read.csv("output/limma/limma_merged.csv", row.names = 1)
ileum_data

dim(ileum_data)
ileum_data <- ileum_data[,colnames(ileum_data)%in%metadata$Sample]
```

#chunk7: making matrix, giving condition information
```{r}
group <- factor(metadata$Condition)
design <- model.matrix(~0 + group)
colnames(design) <- levels(group)
```

#chunk8: fitting data, improving statistical reliability, and running DEG analysis
```{r}
fit <- lmFit(ileum_data, design)
contrast.matrix <- makeContrasts(DiseaseUntreated_vs_Healthy = UC_No - Healthy, levels = design)
fit2 <- contrasts.fit(fit, contrast.matrix)
fit2 <- eBayes(fit2)
#ebayes -improves statistical reliability significantly. 
top_genes <- topTable(fit2, adjust="fdr", number=Inf)
top_genes
write.csv(top_genes, "output/limma/DEGs/DEGsList.csv")

DEGs <- top_genes[top_genes$adj.P.Val < 0.05, ]
total_DEGs <- nrow(DEGs)
upregulated <- sum(DEGs$logFC > 0)
downregulated <- sum(DEGs$logFC < 0)
cat("At FDR < 0.05 - Total DEGs:", total_DEGs, "; Upregulated:", upregulated, "; Downregulated:", downregulated, "\n")
```

#chunk9: Just a PCA plot (Assess the variance explained by the principal components and evaluate whether samples from the same biological group cluster separately from other groups. )
```{r}
library(factoextra)
pca_res <- prcomp(t(ileum_data), scale. = F)
qq <- fviz_pca_ind(pca_res, 
                  label = "metadata$Batch",              
                  habillage = metadata$Condition,     
                  addEllipses = FALSE, 
                  mean.point = FALSE,
                  title = "UC vs Healthy",
                  pointsize = 3                       
) + 
  theme_minimal(base_size = 15) +
     theme(plot.title = element_text(hjust = 0.5, face = "plain", size = 11),
           axis.title = element_text(face = "plain", size = 10),
           axis.text = element_text(face = "plain", size = 9),
           legend.title = element_text(face = "plain", size = 10),
           legend.text = element_text(face = "plain", size = 9))
png("output/limma/PCA/contrastPCA.png",width = 7, height = 5, units = 'in', res = 300)
plot(qq)
dev.off()

```

#chunk10: To visualise genes that found to be differentely expressed using volcano plot.
```{r}
library(ggplot2)
library(ggrepel)

# Genes to highlight (from paper or of special interest)
highlightgenes <- c("REG1A", "FDFT1", "SIK1", "DMBT1", "GGACT", "DUSP21")

# Assign legend category
top_genes$Legend_Category <- "Not significant"
top_genes$Legend_Category[top_genes$adj.P.Val < 0.05] <- "Significant"
top_genes$Legend_Category[rownames(top_genes) %in% highlightgenes &
                         top_genes$adj.P.Val < 0.05] <- "Reported in Papers"
# Color palette
legend_colors <- c("Reported in Papers" = "red",
                   "Significant" = "#1e90ff",
                   "Not significant" = "grey")

# Label subset
label_df <- top_genes[rownames(top_genes) %in% highlightgenes &
                      top_genes$adj.P.Val < 0.05, ]


# Reorder so "Highlighted in Paper" come last
top_genes$Legend_Category <- factor(top_genes$Legend_Category,
                                    levels = c("Not significant", "Significant", "Reported in Papers"))

# Volcano plot
a <- ggplot(top_genes, aes(x = logFC, y = -log10(adj.P.Val),color = Legend_Category)) +
  geom_point(alpha = 0.7, size = 2.5) +
  scale_color_manual(values = legend_colors) +
  geom_vline(xintercept = c(0, 0), linetype = "dashed", color = "grey") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  geom_text_repel(data = label_df,
                  aes(label = rownames(label_df)),
                  color = "black", size = 2.5,
                  box.padding = 0.4, point.padding = 0.3,
                  max.overlaps = 20, show.legend = FALSE) +
#  scale_x_continuous(limits = c(-3, 5)) +
  theme_minimal() +
  labs(title = "UC vs Healthy",
       x = "log2 Fold Change",
       y = "-log10 Adjusted P value",
       color = "Legend") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 11),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 9)
  )

ggsave(filename = "output/limma/Volcano_plots/VolcanoPlot.png", plot = a, width = 8, height = 6, dpi = 300)
```

#chunk11: Pathway analysis: Just on a DEG list: No padj, or FC cutoff: GSEA (we used it here)
# ORA analysis: performed on significant genes (with cut offs) - OLD one - can be used when we have a long list of significant genes.
```{r}
# A data frame 'top_genes' with rownames = gene symbols, and a 'logFC' column
res <- top_genes

# Step 1: Map gene symbols to ENTREZ IDs
ncbi_list <- clusterProfiler::bitr(
  rownames(res), 
  fromType = "SYMBOL", 
  toType = "ENTREZID", 
  OrgDb = org.Hs.eg.db
)

res$names <- rownames(res)
res_mapped <- res %>%
  left_join(ncbi_list, by = c("names" = "SYMBOL")) %>%
  filter(!is.na(ENTREZID)) %>%
  distinct(ENTREZID, .keep_all = TRUE)  # remove duplicates

# Step 2: Create ranked gene list (logFC)
ngenes <- res_mapped$logFC
names(ngenes) <- res_mapped$ENTREZID
ngenes <- sort(ngenes, decreasing = TRUE)
library(BiocParallel)

register(
  SerialParam()
) #for sequencential processing
# === Reactome GSEA === #
enp_gsea <- gsePathway(
  geneList = ngenes,
  organism = "human",
  verbose = FALSE
)

enp_gsea <- setReadable(enp_gsea, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")
pathways <- enp_gsea@result %>% arrange(p.adjust)

top_pathways <- pathways %>% arrange(desc(abs(NES)))

# Plot Top 20 Reactome Pathways
top_pathways$Description <- str_wrap(top_pathways$Description, width = 50)
top_pathways <- top_pathways[1:20, ] %>%
  mutate(Description = fct_reorder(Description, NES))

p <- ggplot(top_pathways, aes(x = NES, y = Description, color = p.adjust, size = setSize)) +
  geom_point(alpha = 0.9) +
  scale_color_gradient(low = "#0072B2", high = "#D55E00", name = "FDR (p.adjust)") +
  scale_size(range = c(3, 10), name = "Gene Set Size") +
  labs(
    title = "Top 20 Reactome Pathways",
    x = "Normalized Enrichment Score",
    y = NULL
    ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.y = element_text(size = 10, color = "black", face = "plain"),
    axis.text.x = element_text(size = 11),
    plot.title = element_text(size = 15, face = "plain", hjust = 0.5),
    plot.subtitle = element_text(size = 13, hjust = 0.5),
    plot.caption = element_text(size = 10, face = "italic", hjust = 0.5),
    legend.title = element_text(size = 12, face = "plain"),
    legend.text = element_text(size = 11)
  )

ggsave("output/limma/pathways/pathwaysplot.png", plot = p, width = 8, height = 6)
write.csv(pathways, "output/limma/pathways/pathways.csv")
```

#chunk12: Now, the same analysis using DESeq2 using raw counts.
```{r}
# Preferred in most cases of having rawcounts: Most important to note that
metadata <- read.csv("E:/seq_bulk/1_SRR_to_IGV_bash_script/metadata_subset.csv")
merged <- read.csv("E:/seq_bulk/1_SRR_to_IGV_bash_script/ileum_raw_counts.csv", check.names = FALSE, row.names = 1)

dim(metadata)
dim(merged)

# Ensure gene names are clean
rownames(merged) <- make.unique(rownames(merged))
pc_df <- read.delim(
  "E:/seq_bulk/2_gtf_to_protein_coding_genes/pc_gene.txt",
  header = TRUE,
  sep = "\t",
  stringsAsFactors = FALSE
)
protein_coding_genes <- unique(
  pc_df$gene_name
)
merged <- merged[
  rownames(merged) %in% protein_coding_genes,
]

dim(metadata)
dim(merged)

library(DESeq2)

# Remove confounded datasets
# -----------------------------
metadata <- metadata[!metadata$Batch %in% c("GSE183620", "GSE193677"),]

# Keep only matching samples in count matrix & Reorder metadata to match count matrix columns
merged <- merged[, colnames(merged) %in% metadata$Sample]
metadata <- metadata[match(colnames(merged), metadata$Sample), ]
all(metadata$Sample == colnames(merged))

# Define condition as factor: Set Healthy as reference group
metadata$Condition <- factor(metadata$Condition)
metadata$Condition <- relevel(metadata$Condition, ref = "Healthy")

# Define batch as factor
metadata$Batch <- factor(metadata$Batch)

# Create DESeq2 dataset
dds <- DESeqDataSetFromMatrix(
  countData = round(merged),
  colData = metadata,
  design = ~ Batch + Condition
)

# Remove very low-count genes (harmless cutoff - optional )
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep, ]

# Run DESeq2 differential expression analysis
dds <- DESeq(dds)

res <- results(dds, contrast = c("Condition", "UC_No", "Healthy"))
res <- res[order(res$padj), ]
res

res_df <- as.data.frame(res)
res_df

write.csv(res_df, "output/DESeq2_output/DESeq2_UC_vs_Healthy_results.csv")

upregulated_genes <- sum(
  res_df$log2FoldChange > 0 &
  res_df$padj < 0.05,
  na.rm = TRUE
)

downregulated_genes <- sum(
  res_df$log2FoldChange < 0 &
  res_df$padj < 0.05,
  na.rm = TRUE
)

cat("Upregulated genes:", upregulated_genes, "\n")
cat("Downregulated genes:", downregulated_genes, "\n")
```
#chunk13
```{r}
# VOLCANO PLOT_DESeq2
library(ggplot2)
library(ggrepel)
# ADD GENE NAMES
res_df$gene <- rownames(res_df)
# DEFINE SIGNIFICANCE
res_df$Significance <- "Not Significant"

res_df$Significance[
  res_df$padj < 0.05 &
  res_df$log2FoldChange > 1
] <- "Upregulated"

res_df$Significance[
  res_df$padj < 0.05 &
  res_df$log2FoldChange < -1
] <- "Downregulated"

############################################################
# GENES TO HIGHLIGHT
############################################################

highlightgenes <- c(
  "REG1A",
  "FDFT1",
  "SIK1",
  "DMBT1",
  "GGACT",
  "DUSP21"
)

pdf(
  "output/DESeq2_output/Volcano_UC_vs_Healthy.pdf",
  width = 8,
  height = 6
)

ggplot(
  res_df,
  aes(
    x = log2FoldChange,
    y = -log10(padj)
  )
) +

geom_point(
  aes(color = Significance),
  alpha = 0.7,
  size = 2
) +

scale_color_manual(
  values = c(
    "Upregulated" = "red",
    "Downregulated" = "blue",
    "Not Significant" = "grey70"
  )
) +

geom_text_repel(
  data = subset(
    res_df,
    gene %in% highlightgenes
  ),
  
  aes(label = gene),
  
  size = 4,
  
  max.overlaps = Inf
) +

theme_classic(base_size = 14) +

labs(
  title = "DESeq2: UC vs Healthy",
  x = "Log2 Fold Change",
  y = "-Log10 Adjusted P-value"
)

dev.off()
```
```{r}
# OVERLAP BETWEEN LIMMA AND DESeq2 DEGs
library(VennDiagram)
# LIMMA SIGNIFICANT GENES
limma_sig <- rownames(
  top_genes[
    top_genes$adj.P.Val < 0.05,
  ]
)

length(limma_sig)
# DESeq2 SIGNIFICANT GENES
deseq_sig <- rownames(
  res_df[
    res_df$padj < 0.05,
  ]
)

length(deseq_sig)

# OVERLAP
common_genes <- intersect(
  limma_sig,
  deseq_sig
)

length(common_genes)
# UNIQUE GENES
limma_unique <- setdiff(
  limma_sig,
  deseq_sig
)

deseq_unique <- setdiff(
  deseq_sig,
  limma_sig
)
# PRINT COUNTS

cat(
  "Limma significant genes:",
  length(limma_sig),
  "\n"
)

cat(
  "DESeq2 significant genes:",
  length(deseq_sig),
  "\n"
)

cat(
  "Common significant genes:",
  length(common_genes),
  "\n"
)

# SAVE OVERLAPPING GENES
write.csv(
  common_genes,
  "output/Common_DEGs_limma_DESeq2.csv",
  row.names = FALSE
)

# VENN DIAGRAM


png(
  "output/Common_DEGs_limma_DESeq2_Venn.png",
  width = 6,
  height = 6,
  units = "in",
  res = 300
)

venn.plot <- venn.diagram(
  
  x = list(
    Limma = limma_sig,
    DESeq2 = deseq_sig
  ),
  
  filename = NULL,
  
  fill = c(
    "skyblue",
    "pink"
  ),
  
  alpha = 0.5,
  
  cex = 1.5,
  
  cat.cex = 1.5,
  
  cat.pos = c(-20, 20)
)

grid::grid.draw(venn.plot)

dev.off()
```

#chunk14
```{r}
############################################################
# LIMMA-VOOM DIFFERENTIAL EXPRESSION ANALYSIS
# (Recommended for RNA-seq raw counts)
############################################################

library(limma)
library(edgeR)

############################################################
# LOAD RAW COUNTS + METADATA
############################################################

metadata <- read.csv(
  "E:/seq_bulk/1_SRR_to_IGV_bash_script/metadata_subset.csv"
)

merged <- read.csv(
  "E:/seq_bulk/1_SRR_to_IGV_bash_script/ileum_raw_counts.csv",
  check.names = FALSE,
  row.names = 1
)

############################################################
# CLEAN GENE NAMES
############################################################

rownames(merged) <- make.unique(
  rownames(merged)
)

############################################################
# KEEP ONLY PROTEIN CODING GENES
############################################################

pc_df <- read.delim(
  "E:/seq_bulk/2_gtf_to_protein_coding_genes/pc_gene.txt",
  header = TRUE,
  sep = "\t",
  stringsAsFactors = FALSE
)

protein_coding_genes <- unique(
  pc_df$gene_name
)

merged <- merged[
  rownames(merged) %in% protein_coding_genes,
]

############################################################
# REMOVE CONFOUNDED DATASETS
############################################################

metadata <- metadata[
  !metadata$Batch %in% c(
    "GSE183620",
    "GSE193677"
  ),
]

############################################################
# MATCH METADATA WITH COUNTS
############################################################

merged <- merged[
  ,
  colnames(merged) %in% metadata$Sample
]

metadata <- metadata[
  match(
    colnames(merged),
    metadata$Sample
  ),
]

all(
  metadata$Sample == colnames(merged)
)

############################################################
# DEFINE FACTORS
############################################################

metadata$Condition <- factor(
  metadata$Condition
)

metadata$Condition <- relevel(
  metadata$Condition,
  ref = "Healthy"
)

metadata$Batch <- factor(
  metadata$Batch
)

############################################################
# CREATE DGEList
############################################################

dge <- DGEList(
  counts = merged
)

############################################################
# FILTER LOWLY EXPRESSED GENES
############################################################

keep <- filterByExpr(
  dge,
  group = metadata$Condition
)

dge <- dge[keep, , keep.lib.sizes = FALSE]

############################################################
# NORMALIZATION
############################################################

dge <- calcNormFactors(dge)

############################################################
# DESIGN MATRIX
############################################################

design <- model.matrix(
  ~ Batch + Condition,
  data = metadata
)

############################################################
# VOOM TRANSFORMATION
############################################################

v <- voom(
  dge,
  design,
  plot = TRUE
)

############################################################
# LIMMA FITTING
############################################################

fit <- lmFit(
  v,
  design
)

fit <- eBayes(
  fit
)

############################################################
# EXTRACT RESULTS
############################################################

top_genes <- topTable(
  fit,
  coef = "ConditionUC_No",
  number = Inf,
  adjust.method = "fdr"
)

############################################################
# SAVE RESULTS
############################################################

write.csv(
  top_genes,
  "E:/seq_bulk/output/limma_voom/DEGs_limma_voom.csv"
)

############################################################
# SIGNIFICANT DEGs
############################################################

DEGs <- top_genes[
  top_genes$P.Value < 0.01,
]

############################################################
# SUMMARY
############################################################

cat(
  "Total significant genes:",
  nrow(DEGs),
  "\n"
)

cat(
  "Upregulated:",
  sum(DEGs$logFC > 0),
  "\n"
)

cat(
  "Downregulated:",
  sum(DEGs$logFC < 0),
  "\n"
)
```


